---
title: "analysis"
author: "Eirik Tveit Solheim"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)
```


## Info

This R markdown document contains the code to reproduce the main results, figures, supplementary figures and supplementary tables presented in the paper.   


## Load required packages

```{r}
library(rmarkdown)
library(reshape2)
library(VennDiagram)
library(ggvenn)
library(cowplot)
library(UpSetR)            
library(WebGestaltR)     
library(fgsea)      
library(circlize)
library(ComplexHeatmap) 
library(openxlsx)
library(readxl)
library(plotrix)
library(ggpubr)            
library(tidyverse)         
```


## Load preprocessed data

Data was generated in `preprocess.Rmd` and `GSEA.Rmd`.  

```{r}
# Sample metadata
metadataTranscriptome <- read.delim("./rawData/metadataTranscriptome.txt", header = TRUE, sep = "\t")
metadataProteome <- read.delim("./rawData/metadataProteome.txt", header = TRUE, sep = "\t")
metadataSecretome <- read.delim("./rawData/metadataSecretome.txt", header = TRUE, sep = "\t")

# Expression data
CPM <- readRDS("./rdsData/CPM.rds")
countMatrixFiltered <- readRDS("./rdsData/countMatrixFiltered.rds")

proteins_filt <- readRDS("./rdsData/proteins_filt.rds")
proteins_keep <- readRDS("./rdsData/proteins_keep.rds")
m.RF <- readRDS("./rdsData/m.RF.rds")

proteins_CCM_filt <- readRDS("./rdsData/proteins_CCM_filt.rds")
proteins_CCM_keep <- readRDS("./rdsData/proteins_CCM_keep.rds")
m_CCM.RF <- readRDS("./rdsData/m_CCM.RF.rds")

# Differential expression results
resTranscriptomeCDR2L <- readRDS("./results/resTranscriptomeCDR2L.rds")
resTranscriptomeCDR2 <- readRDS("./results/resTranscriptomeCDR2.rds")
resTranscriptomeCDR1 <- readRDS("./results/resTranscriptomeCDR1.rds")

resProteomeCDR2L <- readRDS("./results/resProteomeCDR2L.rds")
resProteomeCDR2 <- readRDS("./results/resProteomeCDR2.rds")
resProteomeCDR1 <- readRDS("./results/resProteomeCDR1.rds")

resSecretomeCDR2L <- readRDS("./results/resSecretomeCDR2L.rds")
resSecretomeCDR2 <- readRDS("./results/resSecretomeCDR2.rds")
resSecretomeCDR1 <- readRDS("./results/resSecretomeCDR1.rds")

# Gene set enrichment analysis results
res.gsea.CDR2L.transcriptome <- readRDS("./results/res.gsea.CDR2L.transcriptome.rds")
mainPathways.CDR2L.transcriptome <- readRDS("./results/mainPathways.CDR2L.transcriptome.rds")

res.gsea.CDR2L.proteome <- readRDS("./results/res.gsea.CDR2L.proteome.rds")
mainPathways.CDR2L.proteome <- readRDS("./results/mainPathways.CDR2L.proteome.rds")

res.gsea.CDR2L.secretome <- readRDS("./results/res.gsea.CDR2L.secretome.rds")
mainPathways.CDR2L.secretome <- readRDS("./results/mainPathways.CDR2L.secretome.rds")

HeatPathway <- readRDS("./rdsData/HeatPathway.rds")
```


## Colors

```{r}
colsDatasets <- c("#0073C2FF", "lightblue", "#CD534CFF")
names(colsDatasets) <- c("Proteome", "Secretome", "Transcriptome")

colsCells <- c("#046094", "#DC98FD", "#60B478", "#E47878")
names(colsCells) <- c("CDR2L", "CDR2", "CDR1", "WT")
```


# Figures 

## Figure 1A: Venn diagram

```{r}
venn.diagram(list(Proteome = proteins_keep$Accession, 
                  Secretome = proteins_CCM_keep$Accession,
                  Transcriptome = CPM$uniprotswissprot),
             na = "remove",                                            
             fill = c("#0073C2FF", "lightblue", "#CD534CFF"), 
             alpha = c(0.5, 0.5, 0.5), 
             lwd = 0, 
             cat.cex = 1.5,
             cat.dist = c(0.055, 0.055, 0.085),
             cex = 1.5,
             imagetype = "png",
             filename = "./figs/fig_1A.png")
```


## Figure 1B: Proportion of differentially expressed genes

```{r, fig.height=8}
data.frame(CDR2 = c(resTranscriptomeCDR2 %>% filter(padj < 0.05) %>% nrow() %>% `/` (nrow(resTranscriptomeCDR2)) %>% `*` (100),
                    resProteomeCDR2 %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resProteomeCDR2)) %>% `*` (100),
                    resSecretomeCDR2 %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resSecretomeCDR2)) %>% `*` (100)),
           CDR2L = c(resTranscriptomeCDR2L %>% filter(padj < 0.05) %>% nrow() %>% `/` (nrow(resTranscriptomeCDR2L)) %>% `*` (100),
                     resProteomeCDR2L %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resProteomeCDR2L)) %>% `*` (100),
                     resSecretomeCDR2L %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resSecretomeCDR2L)) %>% `*` (100)),
           CDR1 = c(resTranscriptomeCDR1 %>% filter(padj < 0.05) %>% nrow() %>% `/` (nrow(resTranscriptomeCDR1)) %>% `*` (100),
                    resProteomeCDR1 %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resProteomeCDR1)) %>% `*` (100),
                    resSecretomeCDR1 %>% filter(adj.P.Val < 0.05) %>% nrow() %>% `/` (nrow(resSecretomeCDR1)) %>% `*` (100)),
           set = c("Transcriptome", "Proteome", "Secretome")) %>% 
  reshape2::melt(.) %>% 
  mutate(set = factor(set, levels = c("Transcriptome", "Proteome", "Secretome"))) %>% 
  ggplot(., aes(x = set, y = value)) +
  geom_point(aes(col = variable), size = 3.5) +
  geom_path(aes(group = variable, col = variable), linewidth = 1.8) +
  ylab("% differentially expressed genes") +
  xlab("") +
  scale_color_manual(values=colsCells) +               
  theme_cowplot() +
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15))

ggsave("./figs/fig_1B.png",
       width = 2560, 
       height = 1440, 
       units = "px")
```


## Figure 1C: UpSet plots

```{r, fig.height=8}
# Transcriptome
UpSetTranscriptome <- list(
  CDR2L = resTranscriptomeCDR2L %>% filter(padj < 0.05) %>% pull(gene_name),
  CDR2 = resTranscriptomeCDR2 %>% filter(padj < 0.05) %>% pull(gene_name),
  CDR1 = resTranscriptomeCDR1 %>% filter(padj < 0.05) %>% pull(gene_name))

png(file = "./figs/fig_1C_transcriptome.png", width = 4800, height = 3000, units = "px", res = 600)
UpSetR::upset(fromList(UpSetTranscriptome), sets.bar.color = "#56B4E9",
              order.by = "freq", empty.intersections = "on",
              queries = list(list(query = intersects, params = list("CDR2L"), color = "#046094", active = T),             
                             list(query = intersects, params = list("CDR2L", "CDR2", "CDR1"), color = "orange", active = T)),
              sets.x.label = "# DE genes",
              mainbar.y.label = "Gene Intersections",
              point.size = 4.0, line.size = 1.5,
              text.scale = c(2, 1.5, 1.2, 1.75, 2.5, 2.75))
dev.off()

# Proteome
UpSetProteome <- list(
  CDR2L = resProteomeCDR2L %>% filter(adj.P.Val < 0.05) %>% pull(Gene),
  CDR2 = resProteomeCDR2 %>% filter(adj.P.Val < 0.05) %>% pull(Gene),
  CDR1 = resProteomeCDR1 %>% filter(adj.P.Val < 0.05) %>% pull(Gene))

png(file = "./figs/fig_1C_proteome.png", width = 4800, height = 3000, units = "px", res = 600)
UpSetR::upset(fromList(UpSetProteome), sets.bar.color = "#56B4E9", sets = c("CDR2L", "CDR2", "CDR1"),
              order.by = "freq", empty.intersections = "on",
              queries = list(list(query = intersects, params = list("CDR2L"), color = "#046094", active = T),              
                             list(query = intersects, params = list("CDR2L", "CDR2", "CDR1"), color = "orange", active = T)),
              sets.x.label = "# DE genes",
              mainbar.y.label = "Gene Intersections",
              point.size = 4.0, line.size = 1.5,
              text.scale = c(2, 1.5, 1.2, 1.75, 2.5, 2.75))
dev.off()

# Secretome
UpSetSecretome <- list(
  CDR2L = resSecretomeCDR2L %>% filter(adj.P.Val < 0.05) %>% pull(Gene),
  CDR2 = resSecretomeCDR2 %>% filter(adj.P.Val < 0.05) %>% pull(Gene),
  CDR1 = resSecretomeCDR1 %>% filter(adj.P.Val < 0.05) %>% pull(Gene))

png(file = "./figs/fig_1C_secretome.png", width = 4800, height = 3000, units = "px", res = 600)
UpSetR::upset(fromList(UpSetSecretome), sets.bar.color = "#56B4E9",
              order.by = "freq", empty.intersections = "on",
              queries = list(list(query = intersects, params = list("CDR2L"), color = "#046094", active = T),           
                             list(query = intersects, params = list("CDR2L", "CDR2", "CDR1"), color = "orange", active = T)),
              sets.x.label = "# DE genes",
              mainbar.y.label = "Gene Intersections",
              point.size = 4.0, line.size = 1.5,
              text.scale = c(2, 1.5, 1.2, 1.75, 2.5, 2.75))
dev.off()
```


## Figure 1D: PCA plots

```{r}
# Transcriptome
pcaTranscriptome <- countMatrixFiltered %>% 
  select(1:12) %>% 
  t() %>% 
  prcomp(., scale = TRUE, center = TRUE)

pcaTranscriptome2 <- pcaTranscriptome$x %>% 
  as.data.frame() %>% 
  mutate(sample_id = rownames(.)) %>%                   
  left_join(., metadataTranscriptome, by = "sample_id") %>% 
  # rename
  mutate(condition = case_when(
  condition == "WT" ~ "WT",
  condition == "KO.CDR1" ~ "CDR1",
  condition == "KO.CDR2" ~ "CDR2",
  condition == "KO.CDR2L" ~ "CDR2L")) %>% 
  mutate(condition = fct_relevel(condition, "CDR1", "CDR2", "CDR2L", "WT")) %>%
  mutate(my_condition = ifelse(condition == "CDR2L", 1, 0))

pcaTranscriptome3 <- summary(pcaTranscriptome)
percVarTranscriptome <- round((pcaTranscriptome3$importance[2, c(1:5)] * 100), 1)

lm(PC1 ~ my_condition, pcaTranscriptome2) %>% summary()

ggplot(pcaTranscriptome2, aes(x = PC1, y = PC2, col = condition)) +
  geom_point(size = 6) +
  xlab(paste0("PC1: ", percVarTranscriptome[1], "% variance")) +
  ylab(paste0("PC2: ", percVarTranscriptome[2], "% variance")) +
  theme_cowplot() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 20)) +
  scale_color_manual(values=colsCells)               
  

ggsave("./figs/fig_1D_transcriptome.png",
       width = 2240, 
       height = 1440, 
       units = "px")


# Proteome
pcaProteome <- prcomp(m.RF, scale = TRUE, center = TRUE)

pcaProteome2 <- pcaProteome$x %>% 
  as.data.frame() %>% 
  mutate(sample_id = rownames(.)) %>%                   
  left_join(., metadataProteome, by = "sample_id") %>% 
  mutate(condition = case_when(
  condition == "WT" ~ "WT",
  condition == "KO.CDR1" ~ "CDR1",
  condition == "KO.CDR2" ~ "CDR2",
  condition == "KO.CDR2L" ~ "CDR2L")) %>% 
  mutate(condition = fct_relevel(condition, "CDR1", "CDR2", "CDR2L", "WT")) %>%
  mutate(my_condition = ifelse(condition == "CDR2L", 1, 0))

pcaProteome3 <- summary(pcaProteome)
percVarProteome <- round((pcaProteome3$importance[2, c(1:5)] * 100), 1)

lm(PC1 ~ my_condition, pcaProteome2) %>% summary()

ggplot(pcaProteome2, aes(x = PC1, y = PC2, col = condition)) +
  geom_point(size = 6) +
  xlab(paste0("PC1: ", percVarProteome[1], "% variance")) +
  ylab(paste0("PC2: ", percVarProteome[2], "% variance")) +
  theme_cowplot() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 20)) +
  scale_color_manual(values=colsCells)

ggsave("./figs/fig_1D_proteome.png",
       width = 2240, 
       height = 1440, 
       units = "px")


# Secretome
pcaSecretome <- prcomp(m_CCM.RF, scale = TRUE, center = TRUE)

pcaSecretome2 <- pcaSecretome$x %>% 
  as.data.frame() %>% 
  mutate(sample_id = rownames(.)) %>%                   
  left_join(., metadataSecretome, by = "sample_id") %>% 
  mutate(condition = case_when(
  condition == "WT" ~ "WT",
  condition == "KO.CDR1" ~ "CDR1",
  condition == "KO.CDR2" ~ "CDR2",
  condition == "KO.CDR2L" ~ "CDR2L")) %>% 
  mutate(condition = fct_relevel(condition, "CDR1", "CDR2", "CDR2L", "WT")) %>%
  mutate(my_condition = ifelse(condition == "CDR2L", 1, 0))

pcaSecretome3 <- summary(pcaSecretome)
percVarSecretome <- round((pcaSecretome3$importance[2, c(1:5)] * 100), 1)

lm(PC1 ~ my_condition, pcaSecretome2) %>% summary()

ggplot(pcaSecretome2, aes(x = PC1, y = PC2, col = condition)) +
  geom_point(size = 6) +
  xlab(paste0("PC1: ", percVarSecretome[1], "% variance")) +
  ylab(paste0("PC2: ", percVarSecretome[2], "% variance")) +
  theme_cowplot() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 20)) +
  scale_color_manual(values=colsCells)

ggsave("./figs/fig_1D_secretome.png",
       width = 2240, 
       height = 1440, 
       units = "px")
```


## Figure 2A: Transcriptome GSEA bar plot

```{r, fig.height = 10}
# Collapsed gene sets
GSEA.plot.transcriptome <- mainPathways.CDR2L.transcriptome %>%
  as_tibble() %>% 
  mutate(direction = ifelse(NES < 0, "down", "up")) %>% 
  group_by(direction) %>% 
  dplyr::slice(1:15) %>%                                     
  # Limit number of characters
  mutate(pathway = case_when(
    nchar(pathway) > 40 ~ paste0(substr(pathway, 1, 40), "..."),
    TRUE ~ pathway)) %>% 
  dplyr::rename(FDR = padj)

GSEA.plot.transcriptome %>%  
  ggplot(aes(NES, fct_reorder(pathway, NES), fill = direction)) +
  geom_col(aes(alpha = FDR)) +
  geom_text(aes(x = ifelse(NES < 0, 0.05, -0.05), label = pathway), hjust = ifelse(GSEA.plot.transcriptome$NES < 0, 0, 1), size = 2) +
  ylab("") +
  xlab("Normalized Enrichment Score") +
  theme_cowplot() +
  theme(
    legend.position = "bottom",
    legend.key.height= unit(10, 'pt'),
    legend.key.width= unit(10, 'pt'),
    legend.title = element_text(size=10),
    legend.text = element_text(size=10),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_blank(),
    axis.ticks.length.x = unit(0.5, "line")) +
  scale_fill_manual(values = c("navyblue", "orange")) +
  scale_alpha(range = c(1, 0.25)) +
  guides(fill="none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

ggsave("./figs/fig_2A.png",
       width = 3000, 
       height = 4400, 
       units = "px", 
       dpi = 600)
```


## Figure 2B: Proteome GSEA bar plot

```{r, fig.height = 10}
# Collapsed gene sets
GSEA.plot.proteome <- mainPathways.CDR2L.proteome %>%
  as_tibble() %>% 
  mutate(direction = ifelse(NES < 0, "down", "up")) %>% 
  group_by(direction) %>% 
  dplyr::slice(1:15) %>%                                     
  # Limit number of characters
  mutate(pathway = case_when(
    nchar(pathway) > 40 ~ paste0(substr(pathway, 1, 40), "..."),
    TRUE ~ pathway)) %>% 
  dplyr::rename(FDR = padj)

GSEA.plot.proteome %>%  
  ggplot(aes(NES, fct_reorder(pathway, NES), fill = direction)) +
  geom_col(aes(alpha = FDR)) +
  geom_text(aes(x = ifelse(NES < 0, 0.05, -0.05), label = pathway), hjust = ifelse(GSEA.plot.proteome$NES < 0, 0, 1), size = 2) +
  ylab("") +
  xlab("Normalized Enrichment Score") +
  theme_cowplot() +
  theme(
    legend.position = "bottom",
    legend.key.height= unit(10, 'pt'),
    legend.key.width= unit(10, 'pt'),
    legend.title = element_text(size=10),
    legend.text = element_text(size=10),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_blank(),
    axis.ticks.length.x = unit(0.5, "line")) +
  scale_fill_manual(values = c("navyblue", "orange")) +
  scale_alpha(range = c(1, 0.25)) +
  guides(fill="none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

ggsave("./figs/fig_2B.png",
       width = 3000, 
       height = 4400, 
       units = "px", 
       dpi = 600)
```


## Figure 2C: Secretome GSEA bar plot

```{r, fig.height = 10}
# Collapsed gene sets
GSEA.plot.secretome <- mainPathways.CDR2L.secretome %>%
  as_tibble() %>% 
  mutate(direction = ifelse(NES < 0, "down", "up")) %>% 
  group_by(direction) %>% 
  dplyr::slice(1:15) %>%                                     
  # Limit number of characters
  mutate(pathway = case_when(
    nchar(pathway) > 40 ~ paste0(substr(pathway, 1, 40), "..."),
    TRUE ~ pathway)) %>% 
  dplyr::rename(FDR = padj)

GSEA.plot.secretome %>%  
  ggplot(aes(NES, fct_reorder(pathway, NES), fill = direction)) +
  geom_col(aes(alpha = FDR)) +
  geom_text(aes(x = ifelse(NES < 0, 0.05, -0.05), label = pathway), hjust = ifelse(GSEA.plot.secretome$NES < 0, 0, 1), size = 2) +
  ylab("") +
  xlab("Normalized Enrichment Score") +
  theme_cowplot() +
  theme(
    legend.position = "bottom",
    legend.key.height= unit(10, 'pt'),
    legend.key.width= unit(10, 'pt'),
    legend.title = element_text(size=10),
    legend.text = element_text(size=10),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_blank(),
    axis.ticks.length.x = unit(0.5, "line")) +
  scale_fill_manual(values = c("navyblue", "orange")) +
  scale_alpha(range = c(1, 0.25)) +
  guides(fill="none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

ggsave("./figs/fig_2C.png",
       width = 3000, 
       height = 4400, 
       units = "px", 
       dpi = 600)
```


## Figure 2D: Heatmap of enriched gene sets

```{r, fig.height=10, fig.width=14}
# Set colors
col_fun <- circlize::colorRamp2(c(-2.25, 0, 2.25), c("navyblue", "white", "orange"))

# Top annotation
top_ha = ComplexHeatmap::rowAnnotation(Dataset = c("Proteome", "Secretome", "Transcriptome"),
                                       col = list(Dataset = colsDatasets),
                                       show_annotation_name = F,
                                       show_legend = F,
                                       simple_anno_size = unit(1, "mm"))
# Heatmap
ht1 <- HeatPathway %>% 
  t() %>% 
  ComplexHeatmap::Heatmap(.,
                          rect_gp = grid::gpar(col = "black", lwd = 0.5),
                          col = col_fun,
                          cluster_columns = T,
                          column_dend_reorder = T,
                          row_dend_reorder = F,
                          show_column_dend = F,
                          show_row_dend = F,
                          show_column_names = T, 
                          left_annotation = top_ha,
                          row_names_gp = grid::gpar(fontsize = 5),
                          column_names_gp = grid::gpar(fontsize = 4),
                          row_names_side = "left",                
                          na_col = "white",                               
                          column_names_rot = 45,
                          width = nrow(HeatPathway)*unit(3.5, "mm"),
                          height = ncol(HeatPathway)*unit(3.5, "mm"),
                          heatmap_legend_param = list(
                            title = "NES",
                            direction = "vertical",
                            legend_height = unit(4, "mm"),
                            labels_gp = gpar(fontsize = 4),
                            at = c(-2.5, 0, 2.5),
                            title_gp = gpar(fontsize = 4)))# %>% 
  #ComplexHeatmap::draw(., padding = unit(c(2, 20, 2, 2), "mm"))

# Export image
png(file = "./figs/fig_2D.png", width = 4400, height = 3000, units = "px", res = 600)
draw(ht1)
dev.off()
```


## Figure 3: Heatmap of ribosome biogenesis factors

The genes were annotated according to the tables in [Ribosome biogenesis factors — from names to functions](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10068337/).  

```{r}
ribo_annot <- list(
  rDNA_transcription = c("POLR2E", "POLR2F", "POLR2H", "POLR2L", "POLR2K", "POLR1C", "POLR1D", "POLR1A", "POLR1B", "POLR1F", "POLR1H", "POLR1E", "POLR1G",
                         "TBP", "UBF", "RRN3", "TAF1C", "TAF1B", "TAF1A", "TAF1D", "TAF12", "H3C1", "H4C1", "TOP1", "TOP2A", "CK2A1", "CK2A2", "CK2N",
                         "SPT4H", "SPT5H", "PAF1", "CTR9", "WDR61", "RFT1", "LEO1", "SUPT16H", "SSRP1","TTF1", "CTCF", "MYC", "MAX", "NOLC1", "BRD2", "BRD4", 
                         "KAT7", "JADE3", "LYAR", "SIRT7", "FBL", "UTP4", "WDR43", "UTP5", "HEART1", "UTP10", "UTP15", "UTP17", "WDR75"),
  rRNA_modificiation = c("FBL", "NOP56", "NOP58", "SNU13", "U3-55K", "DKC", "NHP2", "NOP10", "GAR1", "TSR3", "EMG1", "NAT10", "WBSCR22", "DIMT1", "NML", 
                        "RRP8", "NSUN5", "NOP2", "ZCCHC4", "METTL5", "TRMT112"),
  rRNA_processing = c("UTP23", "UTP24", "NOB1", "RMRP", "POP1", "RPP38", "RPP29", "POP5", "RPP25", "RPP20", "RPP14", "RPP30", "RPP21", "RPP40", "LAS1L", "XRN2",
                          "NOL12", "DOS3", "EXOSC10", "EXOSC1", "EXOSC2", "EXOSC3", "EXOSC4", "EXOSC5", "EXOSC6", "EXOSC7", "EXOSC8", "EXOSC9", "MTR4", "PAPD5",
                          "PAPD7", "ZCCHC7", "ZCCHC8", "MPP6", "C1D", "RBM7", "REXO5", "REXO2", "REXO1", "CCR4", "PARN", "TUT4", "TUT7", "ISG20L2", "ERI1"),
  chaperons = c("GRWD1", "HEATR3", "AAMP", "BCCIP", "PDCD2", "NPM", "AK", "CINAP", "AROS", "TSR2"),
  small_subunit_processome = c("UTP4", "CIRH1A", "UTP5", "WDR43", "NOL11", "UTP10", "HEATR1", "UTP15", "UTP17", "WDR75", "MYBBP1A", "PWP2", "UTP6", "UTP12", "WDR3", "TBL3",
                               "UTP18", "WDR36", "DDX21", "NOP2", "FBL", "NOP56", "NOP58", "SNU13", "U3-55K", "U3IP2", "MPHOSPH10", "IMP3", "IMP4", "RRP7A", "NOL6", "CK2A1",
                               "CK2A2", "CK2N", "RRP36", "WDR46", "UTP14A", "WDSOF1", "RCL1", "BMS1", "ENP1", "BYSL", "NOP14", "NOC4L", "EMG1", "AATF", "NGDN", "NOL10",
                               "XRN2", "NKRF", "DHX15", "RSL1D1", "UTP11", "NOL7", "TDIF2", "UTP3", "UTP24", "ESF1", "ABT1", "DDX49", "DDX10", "DDX47", "SRFBP1", "PDCD11",
                               "KRR1", "DDX52", "DEF", "C1orf107", "KRI1", "UTP23", "RBM19", "DDX48", "EIF4A3", "NOM1", "CMS1", "NOP9", "UTP20", "NAT10", "C1orf131",
                               "PNO1", "DIM2", "RRP12", "NOB1", "DHX37", "NML", "RRP8", "FAM207A", "C21orf70", "MYBBP1A", "WBSCR22", "NIP7", "FTSJ3"),
  "40S_maturation_nucleoplasmic" = c("TSR1", "NOB1", "CSNK1D", "CSNK1E", "RIOK1", "RIO1", "RIOK2", "RIO2", "LTV1", "PARN", "FAM207A", "C21orf70"),
  nuclear_export = c("XPO1", "CRM1", "PNO1", "DIM2", "RIOK2", "LTV1", "PDCD2L", "RANBP3", "FAM207A", "C21orf70", "RRP12", "NXF1", "NXT1", "NMD3", "PA2G4", "EBP1", "ZNF593",
                     "SRSF1", "RAE1", "XPO5"),
  "40S_maturation_cytoplasmic" = c("ENP1", "BYSL", "LTV1", "TSR1", "DIMT1", "RRP12", "PNO1", "DIM2", "NOB1", "CSNK1D", "CSNK1E", "RIOK1", "RIO1", "RIOK2", "RIO2", "RIOK3", 
                                     "DHX15", "SON", "PINX1", "HBS1", "PELO", "ABCE1", "EIF5B", "USP16", "LRRC47", "EIF1AD"),
  "60S_maturation_nucleolar" = c("PDCD11", "CEBPZ", "NOC2L", "NIR", "URB1", "URB2", "DDX51", "DDX5", "DDX31", "DDX24", "WDR74", "RPF1", "MAK16", "RRP1", "RBM28", "PUM3", "RRP15",
                               "PPAN", "SURF6","METTL18", "GNL2", "NOP16", "GLTSCR2", "NOP53", "FTSJ3", "MRT4", "MRTO4", "N6AMT1", "LLPH", "ZNF593", "PA2G4", "EBP1", "NVL2",
                               "MDN1", "LAS1L", "NOL9", "PES1", "BOP1", "WDR12", "DDX27", "MK167IP", "RLP7", "PWP1", "RBM34", "DDX18", "BRIX1", "EBNA1BP2", 
                               "NIP7", "NOP2", "RRS1", "BXDC1", "DDX55", "PAK1IP1", "DDX54", "GNL3", "RLP24", "RSL24D1", "EIF6", "GTPBP4", "NSA2", "NLE1", "MYBBP1A", "NOC3L"),
  "60S_formation_of_5S_RNP" = c("REX05", "REX02", "REXO1", "TFIIIA", "HEATR3", "RRS1", "BXDC1"),
  "60S_maturation_nucleoplasmic" = c("CCDC86", "SDAD1", "PELP1", "TEX10", "WDR18", "MDN1", "SENP3", "NF45", "ILF2", "NF90", "ILF3", "NMD3"),
  "60S_maturation_cytoplasmic" = c("SPATA5", "SPATA5L", "CINP", "C1orf109", "RLP24", "RSL24D1", "ZNF593", "GNL3", "LLPH", "TMA16", "NSA2", "GTPBP4", "NOG1", "PA2G4", "EBP1",
                                     "ZNF622", "DNAJC21", "HSPA1A", "HSPA1B", "MRT4", "MRTO4", "DUSP12", "EFL1", "EFTUD1", "SBDS", "EIF6", "LSG1", "AAMP", "NMD3"),
  other_ribosome_biogenesis_factors = c("NPM", "NCL", "DDX1", "HECTD1", "USP36", "UBR5", "VPRBP"))
```

Extract all genes annotated to GOBP_RIBOSOME_BIOGENESIS

```{r}
# Load gene sets from Molecular Signatures Database
GOBP_pathways <- gmtPathways("./refData/c5.go.bp.v2022.1.Hs.symbols.gmt")

# All genes annotated to GOBP_RIBOSOME_BIOGENESIS (n = 310)
RIBO_BIOGENESIS <- GOBP_pathways["GOBP_RIBOSOME_BIOGENESIS"] %>% unlist(use.names = F)
```

```{r, fig.height=14, fig.width=9}
# Combine transcriptomic and proteomic data 
CDR2L.prot.RNA <- resProteomeCDR2L %>% 
  inner_join(x = .,
             y = resTranscriptomeCDR2L, 
             by = c("Accession" = "uniprotswissprot")) %>%
  na.omit()

# Genes annotated to GOBP_RIBOSOME_BIOGENESIS significantly differentially expressed in CDR2L proteome and/or transcriptome
RiboBio <- CDR2L.prot.RNA %>% 
  filter(Gene %in% RIBO_BIOGENESIS) %>%             # n = 255
  filter(adj.P.Val < 0.05 | padj < 0.05) %>%        # n = 206
  distinct(Gene, .keep_all = TRUE) %>%             
  select(Gene, logFC, log2FoldChange, adj.P.Val, padj) %>% 
  rename(logFC.Proteome = logFC, logFC.Transcriptome = log2FoldChange, padj.Proteome = adj.P.Val, padj.Transcriptome = padj)

annot_df <- RiboBio %>% 
  mutate(
    Sig.Transcriptome = case_when(
      padj.Transcriptome < 0.05 ~ "Sig. transcriptome"),
    Sig.Proteome = case_when(
      padj.Proteome < 0.05 ~ "Sig. proteome"),
	  SSU_processome = case_when(
	    Gene %in% ribo_annot$small_subunit_processome ~ "SSU_processome"),
	  rDNA_transcription = case_when(
	    Gene %in% ribo_annot$rDNA_transcription ~ "rDNA_transcription"),
	  rRNA_processing = case_when(
	    Gene %in% ribo_annot$rRNA_modificiation ~ "rRNA_processing",
	    Gene %in% ribo_annot$rRNA_processing ~ "rRNA_processing"),
	  "40S_maturation" = case_when(
	    Gene %in% ribo_annot$"40S_maturation_nucleoplasmic" ~ "40S_maturation",
	    Gene %in% ribo_annot$"40S_maturation_cytoplasmic" ~ "40S_maturation"),
	  "60S_maturation" = case_when(
	    Gene %in% ribo_annot$"60S_maturation_nucleolar" ~ "60S_maturation",
	    Gene %in% ribo_annot$"60S_formation_of_5S_RNP" ~ "60S_maturation",
	    Gene %in% ribo_annot$"60S_maturation_nucleoplasmic" ~ "60S_maturation",
	    Gene %in% ribo_annot$"60S_maturation_cytoplasmic" ~ "60S_maturation"),
	  Chaperon = case_when(
	    Gene %in% ribo_annot$chaperons ~ "chaperons"),
	  Nuclear_export = case_when(
	    Gene %in% ribo_annot$nuclear_export ~ "nuclear_export"),
	  Other_biogenesis_factor = case_when(
	    Gene %in% ribo_annot$other_ribosome_biogenesis_factors ~ "other_ribosome_biogenesis_factors"))

# Row annotation
row_ha = ComplexHeatmap::rowAnnotation(
  Sig.Transcriptome = annot_df$Sig.Transcriptome,
  Sig.Proteome = annot_df$Sig.Proteome,
  SSU_processome = annot_df$SSU_processome,
  rDNA_transcription = annot_df$rDNA_transcription,
  rRNA_processing = annot_df$rRNA_processing,
  "40S_maturation" = annot_df$"40S_maturation",
  "60S_maturation" = annot_df$"60S_maturation",
  Ribosomal_protein_subunit = annot_df$Ribosomal_protein_subunit,
  col = list(
    Sig.Transcriptome = c(
      "Sig. transcriptome" = "black"),
    Sig.Proteome = c(
      "Sig. proteome" = "black"),
    SSU_processome = c(
      "SSU_processome" = "#12086f"),
    rDNA_transcription = c(
      "rDNA_transcription" = "#4361ee"),
    rRNA_processing = c(
      "rRNA_processing" = "#4895ef"),
    "40S_maturation" = c(
      "40S_maturation" = "#4cc9f0"),
    "60S_maturation" = c(
      "60S_maturation" = "lightblue")),
  na_col = "transparent",
  show_annotation_name = F,
  show_legend = T,
  gp = gpar(col = "darkgray", lwd = 0.5))

# Column annotation
top_ha = ComplexHeatmap::HeatmapAnnotation(Dataset = c("Proteome", "Transcriptome"),
                                           col = list(Dataset = colsDatasets),
                                           show_annotation_name = F,
                                           show_legend = F,
                                           simple_anno_size = unit(1, "mm"))
# Colors
col_fun2 <- circlize::colorRamp2(c(-1, 0, 1), c("navyblue", "white", "orange"))

# Heatmap
ht2 <- RiboBio %>% 
  column_to_rownames(var = "Gene") %>% 
  select(logFC.Proteome, logFC.Transcriptome) %>% 
  ComplexHeatmap::Heatmap(.,
                          left_annotation = row_ha[,1:2],
                          right_annotation = row_ha[,3:7],
                          top_annotation = top_ha,
                          #rect_gp = grid::gpar(col = "white", lwd = 1),   # white border between cells
                          col = col_fun2,
                          cluster_rows = T,
                          row_dend_reorder = T,
                          show_row_dend = F,
                          show_column_dend = F,
                          show_column_names = F,
                          row_names_gp = grid::gpar(fontsize = 3),
                          width = ncol(RiboBio %>% select(2,3))*unit(15, "mm"),
                          name = "Log2 Fold Change") 

# Export image
png(file = "./figs/fig_3.png", width = 3300, height = 4400, units = "px", res = 600)
draw(ht2)
dev.off()
```


## Figure 4A: Functionally organized networks of enriched GO terms

Create input used to visualize the results of the GSEA as a functionally organized network in Cytoscape.  
Tab-delimited file with the enriched GO term IDs and associated adjusted p-values.  

```{r}
clueGO_proteome <- res.gsea.CDR2L.proteome %>% 
  ungroup() %>% 
  select(GO_Term_ID, padj) %>% 
  na.omit()
write.table(clueGO_proteome, file = "./cytoscape/clueGO_proteome.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```


## Figure 4B: Heatmap of cell cycle-related genes

```{r, fig.height=14}
GenesOfInterest <- c(
  # CDKs
  "CDK2", "CDK4", "CDK6",
  # Cyclins
  "CCNA2", "CCNB1", "CCNB2", 
  # Cyclin-dependent kinase inhibitor
  "CDKN1A", "CDKN1B", "CDKN2A",
  # Cell division cycle proteins
  "CDC27", "CDC20", "CDC23", "CDC7", 
  # G1/2 (MCM complex)
  "MCM4", "MCM5", "MCM7", "MCM3", "MCM2",
  # Transcription factors
  "E2F4", "MYC", "ETS1", "FOSL1", "FOSB", "JUNB",
  # Key regulators
  "RB1", "RBL1", "RBL2", "MKI67", 
  # DNA damage
  "CHEK1", "TOP2A", "TOP2", "TOP3B",
  # G2/M
 "ESPL1", "MAD2L2", "PLK1",
  # chromosomal passenger complex (CPC)
  "AURKB", "AURKA", "INCENP", "CDCA8",
  # centromeric proteins, kinetochore function and chromosome segregation
  "CENPF", "CENPE", "CENPM", "CENPU",
  # Regulator of chromosome structure during mitosis
  "CDCA2", "CDCA5",
  # condensing complex
  "NCAPH", "NCAPG", "NCAPD2", "NCAPD3", "NCAPG2",
  # regulators of cell cycle
  "BUB1B", "BUB1")

# Order of heatmap
heatmapOrder = c(
  # Cyclin-dependent kinase inhibitor
  "CDKN1B", "CDKN2A",
  # CDKs
  "CDK4", "CDK6", 
  # Cyclins
  "CCNA2", "CCNB1", "CCNB2",
  # centromere proteins
  "CENPE", "CENPF", "CENPM", "CENPU",
  # Regulators of the cell cycle
  "AURKB", "BUB1", "BUB1B", "CHEK1", "MKI67", "RB1", "E2F4",
  # MCM complex
  "MCM2", "MCM3", "MCM4", "MCM5", "MCM7",
  # Condensing complex
  "NCAPD2", "NCAPG", "NCAPG2", "NCAPH", 
  # chromosomal passenger complex (CPC)
  "CDCA8", "INCENP",
  # Others
  "TOP2A", "TOP3B", "CDCA5", "ESPL1", "CDC7")

# Set colors
cellCycleCols = c("Proteome" = "#0073C2FF", "Transcriptome" = "#CD534CFF")
col_fun3 <- circlize::colorRamp2(c(-2, 0, 2), c("navyblue", "white", "orange"))

# Top annotation
top_cellCycle = ComplexHeatmap::HeatmapAnnotation(Dataset = c("Proteome", "Transcriptome"),
                                            col = list(Dataset = cellCycleCols),
                                            simple_anno_size = unit(1, "mm"),
                                            show_annotation_name = F,
                                            show_legend = F)

# Heatmap 
ht3 <- resTranscriptomeCDR2L %>% 
  select(uniprotswissprot, gene_name, log2FoldChange, padj) %>% 
  full_join(x = .,
            y = resProteomeCDR2L %>% select(Accession, Gene, logFC, adj.P.Val),
            by = c("uniprotswissprot" = "Accession")) %>% 
  filter(gene_name %in% GenesOfInterest & "Q8N726" != uniprotswissprot) %>% 
  filter(adj.P.Val < 0.05 & padj < 0.05) %>%                                 
  select(gene_name, logFC, log2FoldChange) %>% 
  dplyr::rename(Proteome = logFC, Transcriptome = log2FoldChange) %>%
  arrange(desc(gene_name)) %>% 
  mutate(gene_name = factor(gene_name, levels = gene_name)) %>% 
  column_to_rownames("gene_name") %>% 
  as.matrix() %>% 
  ComplexHeatmap::Heatmap(.,
                          rect_gp = grid::gpar(col = "black", lwd = 0.5),
                          col = col_fun3,
                          row_order = heatmapOrder,
                          cluster_rows = F,
                          row_dend_reorder = T,
                          show_column_dend = FALSE,
                          show_row_dend = FALSE,
                          top_annotation = top_cellCycle,
                          show_column_names = F,                          
                          row_names_gp = grid::gpar(fontsize = 6),        
                          row_names_side = "left",                        
                          na_col = "white",                               
                          width = ncol(.)*unit(5, "mm"),
                          height = nrow(.)*unit(5, "mm"),
                          heatmap_legend_param = list(
                            title = "Log2 Fold Change",
                            #legend_height = unit(4, "mm"),
                            labels_gp = gpar(fontsize = 5),
                            title_gp = gpar(fontsize = 6),
                            title_position = "leftcenter-rot"))   

# Export image
png(file = "./figs/fig_4B.png", width = 2200, height = 4400, units = "px", res = 600)
draw(ht3)
dev.off()
```


## Figure 5A: Venn diagram showing the number of shared and unique differentially expressed genes between transcriptome and proteome

```{r}
venn.diagram(list(Transcriptome = resTranscriptomeCDR2L %>% filter(padj < 0.05) %>% pull(uniprotswissprot),
                  Proteome = resProteomeCDR2L %>% filter(adj.P.Val < 0.05) %>% pull(Accession)),
             na = "remove",                                           
             fill = c("#CD534CFF", "#0073C2FF"), 
             alpha = c(0.5, 0.5), 
             lwd = 0, 
             margin = 0.05,
             cat.cex = 1.5,
             cex = 1.5,
             imagetype = "png",
             filename = "./figs/fig_5A.png")
```


## Figure 5B: Scatter plot of log2 fold changes in transcriptome and proteome

```{r, fig.height=8}
# Combine transcriptomic and proteomic data 
CDR2L.prot.RNA <- resProteomeCDR2L %>% 
  inner_join(x = .,
             y = resTranscriptomeCDR2L, 
             by = c("Accession" = "uniprotswissprot")) %>%
  na.omit() %>%  
  mutate(Groups = case_when(
    log2FoldChange < 0 & logFC < 0 ~ "Down/Down",
    log2FoldChange > 0 & logFC > 0 ~ "Up/Up",
    log2FoldChange < 0 & logFC > 0 ~ "Down transcriptome/Up proteome",
    log2FoldChange > 0 & logFC < 0 ~ "Up transcriptome/Down proteome"))

# Number of genes in each group
CDR2L.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  count(Groups)

# Scatter plot
CDR2L.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  ggplot(aes(x = log2FoldChange, y = logFC)) +
  geom_point(aes(color = Groups), size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  geom_vline(xintercept = 0, lty = 2, color = "lightgrey") +
  geom_hline(yintercept = 0, lty = 2, color = "lightgrey") +
  xlab("Log2FC Transcriptome") +
  ylab("Log2FC Proteome") +
  scale_color_manual(values=c("#C77CFF", "#F8766D", "#7CAE00", "#00BFC4FF")) +
  ggpubr::stat_cor(method = "spearman") +
  theme_cowplot() +
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 15)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave("./figs/fig_5B.png",
       width = 2560, 
       height = 1440, 
       units = "px")
```


## Figure 6: Protein-protein network

Input used to create protein-protein interaction network of genes differentially expressed in both transcriptome and proteome in Cytoscape.

```{r}
CDR2L.prot.RNA.sig.DE <- CDR2L.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>% 
  select(Accession, Gene, logFC, adj.P.Val, log2FoldChange, padj, Groups) %>% 
  dplyr::rename(protein_logFC = logFC, protein_padj = adj.P.Val, gene_logFC = log2FoldChange, gene_padj = padj) %>% 
  mutate_if(is.numeric, ~round(., 4))

write_tsv(CDR2L.prot.RNA.sig.DE, "./cytoscape/CDR2L.prot.RNA.sig.DE.tsv")
```


## Figure 7: Proliferation assay

The proliferation assay was performed using the IncuCyte Live-Cell Analysis System (Sartorius). The raw data was exported and used to create the figure presented in the paper. 

```{r, fig.height=8}
prolif <- read.delim("./data/Proliferation_WT_CDR2L.txt", header = TRUE, sep = "\t", dec = ",", skip = 1)

prolifMean <- prolif %>% 
  mutate(
    WT = rowMeans(.[, 3:18]),
    CDR2L = rowMeans(.[, 19:34])) %>% 
  select(Elapsed, WT, CDR2L) %>% 
  column_to_rownames("Elapsed") %>% 
  t() %>% 
  reshape2::melt(., id.vars = rownames(.)) %>% 
  rename(mean = value, 
         sample = Var1,
         time = Var2)

prolifSE <- prolif %>% 
  mutate(
    WT.se = apply(.[, 3:18], 1, plotrix::std.error),               # use sd to calculate standard deviation
    CDR2L.se = apply(.[, 19:34], 1, plotrix::std.error)) %>% 
  select(Elapsed, WT.se, CDR2L.se) %>% 
  column_to_rownames("Elapsed") %>% 
  t() %>% 
  reshape2::melt(., id.vars = rownames(.)) %>% 
  rename(SE = value,
         sample = Var1,
         time = Var2)

prolifComb <- bind_cols(prolifMean, prolifSE[, 3]) %>% 
  rename(SE = ...4)

# Lineplot
prolifComb %>% 
  ggplot(aes(x = time, y = mean, color = sample)) +
  geom_point(size = 3) +
  geom_path(size = 1) +
  geom_errorbar(aes(ymin = mean-SE, ymax = mean+SE), width = 1) +
  #geom_segment(aes(x = 0, y = 90, xend = 66, yend = 90), color = "gray") +
  #geom_segment(aes(x = 66, y = 90, xend = 66, yend = 0), color = "gray") +
  xlab("Hours") +
  ylab("Confluence (%)") +
  scale_color_manual(breaks = c("CDR2L", "WT"),
                     values=c("#046094", "#E47878")) +
  scale_x_continuous(n.breaks = 20) +
  scale_y_continuous(n.breaks = 20) +
  #ylim(0, 100) +
  #xlim(0, 100) +
  theme_cowplot() +
  theme(legend.title=element_blank(),
        legend.position = c(0.8, 0.8))

ggsave("./figs/fig_7.png",
       width = 2560, 
       height = 1440, 
       units = "px")
```


# Supplementary Figures 

## Supplementary Figure S1: Correlation between datasets

```{r}
# Prepare transcriptome data
cor_transcriptome_avg <- CPM %>% 
  distinct(uniprotswissprot, .keep_all = TRUE) %>%                       
  na.omit(uniprotswissprot) %>%
  remove_rownames() %>% 
  column_to_rownames("uniprotswissprot") %>%
  mutate(
    WT.avg = rowMeans(.[, 3:5]),    
    KO.CDR1.avg = rowMeans(.[, 6:8]),
    KO.CDR2.avg = rowMeans(.[, 9:11]),
    KO.CDR2L.avg = rowMeans(.[, 12:14])) %>% 
  select(15:18) %>%  
  t() %>%
  reshape2::melt(., id.vars = rownames(.)) %>% 
  dplyr::rename(sample = Var1,
         protein = Var2) %>% 
  mutate(type = "transcriptome") 

# Prepare proteome data
cor_proteome_avg <- m.RF %>% 
  log2() %>% 
  t() %>% 
  as.data.frame() %>% 
  mutate(
    WT.avg = rowMeans(.[, 1:3]),    
    KO.CDR1.avg = rowMeans(.[, 4:6]),
    KO.CDR2.avg = rowMeans(.[, 7:9]),
    KO.CDR2L.avg = rowMeans(.[, 10:12])) %>% 
  select(13:16) %>% 
  t() %>% 
  reshape2::melt(., id.vars = rownames(.)) %>% 
  dplyr::rename(sample = Var1,
         protein = Var2) %>% 
  mutate(type = "proteome")

# Prepare secretome data
cor_secretome_avg <- m_CCM.RF %>% 
  log2() %>% 
  t() %>% 
  as.data.frame() %>% 
  mutate(
    WT.avg = rowMeans(.[, 1:3]),    
    KO.CDR1.avg = rowMeans(.[, 4:6]),
    KO.CDR2.avg = rowMeans(.[, 7:9]),
    KO.CDR2L.avg = rowMeans(.[, 10:12])) %>% 
  select(13:16) %>% 
  t() %>% 
  reshape2::melt(., id.vars = rownames(.)) %>% 
  dplyr::rename(sample = Var1,
         protein = Var2) %>% 
  mutate(type = "secretome")
```

```{r, fig.height = 8, fig.width=10}
# Transcriptome vs Proteome
int.prot.RNA <- intersect(proteins_filt$Accession, CPM$uniprotswissprot)       

cor_proteome_transcriptome <- rbind(cor_proteome_avg, cor_transcriptome_avg) %>% filter(protein %in% int.prot.RNA) %>%
  pivot_wider(names_from = "type", values_from = "value")                         

cor_proteome_transcriptome %>% 
  ggplot(aes(x = transcriptome, y = proteome)) +    # RNA (x-axis) as independent variable, protein as dependent variable
  geom_point(color = "#565656", size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  xlab("Transcriptome") +
  ylab("Proteome") +
  ggpubr::stat_cor(method = "spearman") +
  cowplot::theme_cowplot()

ggsave("./figs/fig_S1A.png",
       width = 1800, 
       height = 1440, 
       units = "px")

# Transcriptome vs Secretome
int.CCM.RNA <- intersect(proteins_CCM_filt$Accession, CPM$uniprotswissprot)       

cor_secretome_transcriptome <- rbind(cor_secretome_avg, cor_transcriptome_avg) %>% filter(protein %in% int.CCM.RNA) %>%   
  pivot_wider(names_from = "type", values_from = "value")                              

cor_secretome_transcriptome %>% 
  ggplot(aes(x = transcriptome, y = secretome)) + 
  geom_point(color = "#565656", size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  xlab("Transcriptome") +
  ylab("Secretome") +
  ggpubr::stat_cor(method = "spearman") +
  cowplot::theme_cowplot()

ggsave("./figs/fig_S1B.png",
       width = 1800, 
       height = 1440, 
       units = "px")

## Proteome vs Secretome
int.prot.CCM <- intersect(proteins_filt$Accession, proteins_CCM_filt$Accession)       

cor_proteome_secretome <- rbind(cor_proteome_avg, cor_secretome_avg) %>% filter(protein %in% int.prot.CCM) %>%   
  pivot_wider(names_from = "type", values_from = "value")   

cor_proteome_secretome %>% 
  ggplot(aes(x = proteome, y = secretome)) +    
  geom_point(color = "#565656", size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  xlab("Proteome") +
  ylab("Secretome") +
  ggpubr::stat_cor(method = "spearman") +
  cowplot::theme_cowplot()

ggsave("./figs/fig_S1C.png",
       width = 1800, 
       height = 1440, 
       units = "px")
```


## Supplementary Figure S2: Scatter plots of log2 fold changes in transcriptome and proteome

```{r, fig.height=8}
# Knockout CDR2 cells

# Combine transcriptomic and proteomic data 
CDR2.prot.RNA <- resProteomeCDR2 %>% 
  inner_join(x = .,
             y = resTranscriptomeCDR2, 
             by = c("Accession" = "uniprotswissprot")) %>%
  na.omit() %>%  
  mutate(Groups = case_when(
    log2FoldChange < 0 & logFC < 0 ~ "Down/Down",
    log2FoldChange > 0 & logFC > 0 ~ "Up/Up",
    log2FoldChange < 0 & logFC > 0 ~ "Down transcriptome/Up proteome",
    log2FoldChange > 0 & logFC < 0 ~ "Up transcriptome/Down proteome"))

# Number of genes in each group
CDR2.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  dplyr::count(Groups)

# Scatter plot
CDR2.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  ggplot(aes(x = log2FoldChange, y = logFC)) +
  geom_point(aes(color = Groups), size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  geom_vline(xintercept = 0, lty = 2, color = "lightgrey") +
  geom_hline(yintercept = 0, lty = 2, color = "lightgrey") +
  xlab("Log2FC Transcriptome") +
  ylab("Log2FC Proteome") +
  scale_color_manual(values=c("#C77CFF", "#F8766D", "#7CAE00", "#00BFC4FF")) +
  ggpubr::stat_cor(method = "spearman") +
  cowplot::theme_cowplot() +
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 15)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave("./figs/fig_S2A.png",
       width = 2560, 
       height = 1440, 
       units = "px")


# Knockout CDR1 cells

# Combine transcriptomic and proteomic data 
CDR1.prot.RNA <- resProteomeCDR1 %>% 
  inner_join(x = .,
             y = resTranscriptomeCDR1, 
             by = c("Accession" = "uniprotswissprot")) %>%
  na.omit() %>%  
  mutate(Groups = case_when(
    log2FoldChange < 0 & logFC < 0 ~ "Down/Down",
    log2FoldChange > 0 & logFC > 0 ~ "Up/Up",
    log2FoldChange < 0 & logFC > 0 ~ "Down transcriptome/Up proteome",
    log2FoldChange > 0 & logFC < 0 ~ "Up transcriptome/Down proteome"))

# Number of genes in each group
CDR1.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  dplyr::count(Groups)

# Scatter plot
CDR1.prot.RNA %>% 
  filter(padj < 0.05 & adj.P.Val < 0.05) %>%
  ggplot(aes(x = log2FoldChange, y = logFC)) +
  geom_point(aes(color = Groups), size = 0.75) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE) +
  geom_vline(xintercept = 0, lty = 2, color = "lightgrey") +
  geom_hline(yintercept = 0, lty = 2, color = "lightgrey") +
  xlab("Log2FC Transcriptome") +
  ylab("Log2FC Proteome") +
  scale_color_manual(values=c("#C77CFF", "#F8766D", "#7CAE00", "#00BFC4FF")) +
  ggpubr::stat_cor(method = "spearman") +
  cowplot::theme_cowplot() +
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 15),
        text = element_text(size = 15),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 15)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave("./figs/fig_S2B.png",
       width = 2560, 
       height = 1440, 
       units = "px")
```







# Main Results 

## Section 1: Knockout of genes encoding CDR proteins induces expression changes in the transcriptome, proteome, and secretome of ovarian cancer cells  

### Proportion of common and uniquely differentially expressed genes 

```{r}
# Proportion of common DE genes in each dataset:
print("Proportion of common DEGs in transcriptome")
length(Reduce(intersect, list(UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR1))) %>% `/` (length(unique(c(UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR1))))

print("Proportion of common DEGs in proteome")
length(Reduce(intersect, list(UpSetProteome$CDR2L, UpSetProteome$CDR2, UpSetProteome$CDR1))) %>% `/` (length(unique(c(UpSetProteome$CDR2L, UpSetProteome$CDR2, UpSetProteome$CDR1))))

print("Proportion of common DEGs in secretome")
length(Reduce(intersect, list(UpSetSecretome$CDR2L, UpSetSecretome$CDR2, UpSetSecretome$CDR1))) %>% `/` (length(unique(c(UpSetSecretome$CDR2L, UpSetSecretome$CDR2, UpSetSecretome$CDR1))))

# Proportion of DE genes unique to CDR2L in each dataset:
print("Proportion of DEGs unique to CDR2L in transcriptome")
length(Reduce(setdiff, list(UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR1))) %>% `/` (length(unique(c(UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR1))))

print("Proportion of DEGs unique to CDR2L in proteome")
length(Reduce(setdiff, list(UpSetProteome$CDR2L, UpSetProteome$CDR2, UpSetProteome$CDR1))) %>% `/` (length(unique(c(UpSetProteome$CDR2L, UpSetProteome$CDR2, UpSetProteome$CDR1))))

print("Proportion of DEGs unique to CDR2L in secretome")
length(Reduce(setdiff, list(UpSetSecretome$CDR2L, UpSetSecretome$CDR2, UpSetSecretome$CDR1))) %>% `/` (length(unique(c(UpSetSecretome$CDR2L, UpSetSecretome$CDR2, UpSetSecretome$CDR1))))

# Proportion of DE genes unique to CDR2 in each dataset:
print("Proportion of DEGs unique to CDR2 in transcriptome")
length(Reduce(setdiff, list(UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR1))) %>% `/` (length(unique(c(UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR1))))

print("Proportion of DEGs unique to CDR2 in proteome")
length(Reduce(setdiff, list(UpSetProteome$CDR2, UpSetProteome$CDR2L, UpSetProteome$CDR1))) %>% `/` (length(unique(c(UpSetProteome$CDR2, UpSetProteome$CDR2L, UpSetProteome$CDR1))))

print("Proportion of DEGs unique to CDR2 in secretome")
length(Reduce(setdiff, list(UpSetSecretome$CDR2, UpSetSecretome$CDR2L, UpSetSecretome$CDR1))) %>% `/` (length(unique(c(UpSetSecretome$CDR2, UpSetSecretome$CDR2L, UpSetSecretome$CDR1))))

# Proportion of DE genes unique to CDR1 in each dataset:
print("Proportion of DEGs unique to CDR1 in transcriptome")
length(Reduce(setdiff, list(UpSetTranscriptome$CDR1, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR2L))) %>% `/` (length(unique(c(UpSetTranscriptome$CDR1, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR2L))))

print("Proportion of DEGs unique to CDR1 in proteome")
length(Reduce(setdiff, list(UpSetProteome$CDR1, UpSetProteome$CDR2, UpSetProteome$CDR2L))) %>% `/` (length(unique(c(UpSetProteome$CDR1, UpSetProteome$CDR2, UpSetProteome$CDR2L))))

print("Proportion of DEGs unique to CDR1 in secretome")
length(Reduce(setdiff, list(UpSetSecretome$CDR1, UpSetSecretome$CDR2, UpSetSecretome$CDR2L))) %>% `/` (length(unique(c(UpSetSecretome$CDR1, UpSetSecretome$CDR2, UpSetSecretome$CDR2L))))
```


### Over-representation analysis of genes differentially expressed in all knockout cell lines 
  
Transcriptome: 

```{r}
# Set seed
set.seed(123)

# Intersect of differentially expressed genes
DEGs.transcriptome <- Reduce(intersect, list(UpSetTranscriptome$CDR2L, UpSetTranscriptome$CDR2, UpSetTranscriptome$CDR1))

# Over-representation analysis
enrichDatabase <- c("geneontology_Biological_Process_noRedundant", "geneontology_Molecular_Function_noRedundant", "geneontology_Cellular_Component_noRedundant", "pathway_KEGG")

WG.transcriptome <- lapply(enrichDatabase, function(enrichName){
  message(paste0("Running ", enrichName, "..."))
  WebGestaltR(enrichMethod = "ORA",
              organism = "hsapiens",
              enrichDatabase = enrichName,
              interestGene = DEGs.transcriptome,
              interestGeneType = "genesymbol",
              collapseMethod = "mean",
              referenceGene = CPM$gene_name,
              referenceGeneType = "genesymbol",
              sigMethod = "fdr",
              fdrMethod = "BH",
              minNum = 10,
              maxNum = 500,
              isOutput = FALSE) %>% 
    as_tibble() %>% 
    mutate(Set = enrichName)
})
names(WG.transcriptome) <- names(enrichDatabase)
df.WG.transcriptome <- do.call(rbind, WG.transcriptome)

# Compile results in a single table
res.WG.transcriptome <- df.WG.transcriptome %>%
  arrange(FDR) %>%
  select(geneSet, description, size, overlap, expect, enrichmentRatio, pValue, FDR, Set, userId)
```

Proteome: 

```{r}
# Set seed
set.seed(123)

# Intersect of differentially expressed genes
DEGs.proteome <- Reduce(intersect, list(UpSetProteome$CDR2L, UpSetProteome$CDR2, UpSetProteome$CDR1))

# Over-representation analysis
enrichDatabase <- c("geneontology_Biological_Process_noRedundant", "geneontology_Molecular_Function_noRedundant", "geneontology_Cellular_Component_noRedundant", "pathway_KEGG")

WG.proteome <- lapply(enrichDatabase, function(enrichName){
  message(paste0("Running ", enrichName, "..."))
  WebGestaltR(enrichMethod = "ORA",
              organism = "hsapiens",
              enrichDatabase = enrichName,
              interestGene = DEGs.proteome,
              interestGeneType = "genesymbol",
              collapseMethod = "mean",
              referenceGene = proteins_keep$Gene,
              referenceGeneType = "genesymbol",
              sigMethod = "fdr",
              fdrMethod = "BH",
              minNum = 10,
              maxNum = 500,
              isOutput = FALSE) %>% 
    as_tibble() %>% 
    mutate(Set = enrichName)
})
names(WG.proteome) <- names(enrichDatabase)
df.WG.proteome <- do.call(rbind, WG.proteome)

# Compile results in a single table
res.WG.proteome <- df.WG.proteome %>%
  arrange(FDR) %>%
  select(geneSet, description, size, overlap, expect, enrichmentRatio, pValue, FDR, Set, userId)
```

Secretome: 

```{r}
# Set seed
set.seed(123)

# Intersect of differentially expressed genes
DEGs.secretome <- Reduce(intersect, list(UpSetSecretome$CDR2L, UpSetSecretome$CDR2, UpSetSecretome$CDR1))

# Over-representation analysis
enrichDatabase <- c("geneontology_Biological_Process_noRedundant", "geneontology_Molecular_Function_noRedundant", "geneontology_Cellular_Component_noRedundant", "pathway_KEGG")

WG.secretome <- lapply(enrichDatabase, function(enrichName){
  message(paste0("Running ", enrichName, "..."))
  WebGestaltR(enrichMethod = "ORA",
              organism = "hsapiens",
              enrichDatabase = enrichName,
              interestGene = DEGs.secretome,
              interestGeneType = "genesymbol",
              collapseMethod = "mean",
              referenceGene = proteins_CCM_keep$Gene,
              referenceGeneType = "genesymbol",
              sigMethod = "fdr",
              fdrMethod = "BH",
              minNum = 10,
              maxNum = 500,
              isOutput = FALSE) %>% 
    as_tibble() %>% 
    mutate(Set = enrichName)
})
names(WG.secretome) <- names(enrichDatabase)
df.WG.secretome <- do.call(rbind, WG.secretome)

# Compile results in a single table
res.WG.secretome <- df.WG.secretome %>%
  arrange(FDR) %>%
  select(geneSet, description, size, overlap, expect, enrichmentRatio, pValue, FDR, Set, userId)
```


## Section 2: Knockout of CDR2L results in a unique transcriptomic and proteomic expression profile

Genes driving the separation between CDR2L and the other cell lines in the principal component analysis. 

```{r}
pcaVarTranscriptome <- factoextra::get_pca_var(pcaTranscriptome)
pcaVarProteome <- factoextra::get_pca_var(pcaProteome)

Venn.PCA <- list(
  Transcriptome = as.data.frame(pcaVarTranscriptome$contrib) %>%
    arrange(-Dim.1) %>% 
    rownames_to_column(., "gene_id") %>% 
    left_join(x = .,
              y = CPM %>% select(gene_id, uniprotswissprot),
              by = "gene_id") %>% 
    pull(uniprotswissprot) %>% 
    head(n = 1000),
  Proteome = as.data.frame(pcaVarProteome$contrib) %>% 
    arrange(-Dim.1) %>% 
    rownames(.) %>% 
    head(n = 1000))

ggvenn(Venn.PCA, fill_color = c("#CD534CFF", "#0073C2FF"), stroke_size = 0.5, set_name_size = 4)
```


## Section 5: Knockout of CDR2L causes downregulation of key regulators of the cell cycle

Input for transcription factor enrichment analysis performed using [ChIP-X Enrichment Analysis version 3](https://maayanlab.cloud/chea3/).

```{r}
Input_ChEA3 <- resTranscriptomeCDR2L %>% filter(padj < 0.05) %>% pull(gene_name)
```

Results

```{r}
res_ChEA3 <- read.xlsx("./supplement/supplementary_table_S9.xlsx")
```


## Section 6: CDR2L interacts with ribosome biogenesis factors and regulators of the cell cycle

Intersect of potential interaction partners of CDR2L and ribosome biogenesis factors, eukaryotic initiation factors and cell cycle proteins.  
Potential interaction partners identified by co-immunoprecipitation (co-IP) coupled with mass spectrometry. Data originally produced by [Herdlevaer et al. (2020)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7664253/).

```{r}
IP.hits <- readxl::read_xlsx(path = "./data/Interaction_partners_of_CDR2L.xlsx")

# Intersect with ribosome biogenesis factors
intersect(IP.hits$Gene, annot_df$Gene) %>% sort()

# Intersect with eukaryotic initiation factors
intersect(IP.hits$Gene,
          filter(resProteomeCDR2L, adj.P.Val < 0.05 & grepl("^EIF", Gene)) %>% pull(Gene)) %>% sort()

# Intersect with cell cycle proteins
intersect(IP.hits$Gene, GenesOfInterest) %>% sort()
```





# Supplementary Tables

```{r}
# Supplementary Table S1: Transcriptome dataset
write.xlsx(CPM, file = "./supplement/supplementary_table_S1.xlsx")

# Supplementary Table S2: Proteome dataset
suppTab2 <- proteins_keep %>% 
  select(Accession, Gene, Description, 6:17)
write.xlsx(suppTab2, file = "./supplement/supplementary_table_S2.xlsx")

# Supplementary Table S3: Secretome dataset
suppTab3 <- proteins_CCM_keep %>% 
  select(Accession, Gene, Description, 6:17)
write.xlsx(suppTab3, file = "./supplement/supplementary_table_S3.xlsx")

# Supplementary Table S4: Differential expression analysis of CDR2L-knockout cells
names <- list('Transcriptome' = resTranscriptomeCDR2L, 'Proteome' = resProteomeCDR2L, 'Secretome' = resSecretomeCDR2L)
write.xlsx(names, file = "./supplement/supplementary_table_S4.xlsx")

# Supplementary Table S5: Differential expression analysis of CDR2-knockout cells
names <- list('Transcriptome' = resTranscriptomeCDR2, 'Proteome' = resProteomeCDR2, 'Secretome' = resSecretomeCDR2)
write.xlsx(names, file = "./supplement/supplementary_table_S5.xlsx")

# Supplementary Table S6: Differential expression analysis of CDR1-knockout cells
names <- list('Transcriptome' = resTranscriptomeCDR1, 'Proteome' = resProteomeCDR1, 'Secretome' = resSecretomeCDR1)
write.xlsx(names, file = "./supplement/supplementary_table_S6.xlsx")

# Supplementary Table S7: Over-representation analysis of common genes
names <- list('Transcriptome' = res.WG.transcriptome, 'Proteome' = res.WG.proteome, 'Secretome' = res.WG.secretome)
write.xlsx(names, file = "./supplement/supplementary_table_S7.xlsx")

# Supplementary Table S8: Gene set enrichment analysis of CDR2L-knockout cells 
names <- list('Transcriptome' = res.gsea.CDR2L.transcriptome, 'Proteome' = res.gsea.CDR2L.proteome, 'Secretome' = res.gsea.CDR2L.secretome)
write.xlsx(names, file = "./supplement/supplementary_table_S8.xlsx")
```

